---
title: "Assignment 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Qianmu Zheng"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:**
- Apply spatial operations to answer policy-relevant research questions
- Integrate census demographic data with spatial analysis
- Create publication-quality visualizations and maps
- Work with spatial data from multiple sources
- Communicate findings effectively for policy audiences

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data:
- Pennsylvania county boundaries
- Pennsylvania hospitals (from lecture data)
- Pennsylvania census tracts

**Your Task:**
```{r}
# Load required packages
library(sf)
library(tidyverse)
library(tigris)
library(tidycensus)
library(scales)
library(patchwork)
library(here)
# Load spatial data
data_path <- "D:/Desktop/musa/5080_policy/5080/MUSA-5080-Fall-2025/lectures/week-04/data"
census_api_key("940dffa67b928a0518accaf8839aa7b4762b11ab")

pa_counties <- st_read(file.path(data_path, "Pennsylvania_County_Boundaries.shp"))
hospitals <- st_read(file.path(data_path, "hospitals.geojson"))
census_tracts <- tracts(state = "PA", cb = TRUE)

# Check that all data loaded correctly
hospitals <- st_transform(hospitals, st_crs(pa_counties))
census_tracts <- st_transform(census_tracts, st_crs(pa_counties))

head(pa_counties)
head(hospitals)
head(census_tracts)
```

**Questions to answer:**
- How many hospitals are in your dataset?
- How many census tracts?
- What coordinate reference system is each dataset in?
```{r}
num_hospitals <- nrow(hospitals)
cat("Number of hospitals in dataset:", num_hospitals, "\n")

num_tracts <- nrow(census_tracts)
cat("Number of census tracts in dataset:", num_tracts, "\n")

cat("\n---- Coordinate Reference Systems ----\n")
cat("Counties CRS:\n"); print(st_crs(pa_counties))
cat("\nHospitals CRS:\n"); print(st_crs(hospitals))
cat("\nCensus Tracts CRS:\n"); print(st_crs(census_tracts))
```

---

#### Step 2: Get Demographic Data 

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:**
- Total population
- Median household income
- Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**
```{r}
# Get demographic data from ACS
library(tidycensus)
library(dplyr)
library(sf)

acs_year <- 2022

pa_basic <- get_acs(
  geography = "tract",
  state = "PA",
  variables = c(
    total_pop = "B01003_001",
    median_income = "B19013_001"
  ),
  year = acs_year,
  survey = "acs5",
  geometry = FALSE,
  output = "wide"
)

pa_age <- get_acs(
  geography = "tract",
  state = "PA",
  variables = c(
    male_65_66 = "B01001_020",
    male_67_69 = "B01001_021",
    male_70_74 = "B01001_022",
    male_75_79 = "B01001_023",
    male_80_84 = "B01001_024",
    male_85plus = "B01001_025",
    female_65_66 = "B01001_044",
    female_67_69 = "B01001_045",
    female_70_74 = "B01001_046",
    female_75_79 = "B01001_047",
    female_80_84 = "B01001_048",
    female_85plus = "B01001_049"
  ),
  year = acs_year,
  survey = "acs5",
  geometry = FALSE
)

pa_age_summary <- pa_age %>%
  group_by(GEOID) %>%
  summarize(pop_65_over = sum(estimate, na.rm = TRUE))

pa_demo <- pa_basic %>%
  left_join(pa_age_summary, by = "GEOID") %>%
  select(GEOID, total_popE, median_incomeE, pop_65_over) %>%
  rename(
    total_pop = total_popE,
    median_income = median_incomeE
  )

# Join to tract boundaries
census_tracts_demo <- census_tracts %>%
  left_join(pa_demo, by = c("GEOID" = "GEOID"))


# What year of ACS data?
cat("ACS Year Used:", acs_year, "\n")

# How many tracts have missing income data?
missing_income <- sum(is.na(census_tracts_demo$median_income))
cat("Number of tracts with missing income data:", missing_income, "\n")

# What is the median income across all PA census tracts?
overall_median_income <- median(census_tracts_demo$median_income, na.rm = TRUE)
cat("Median income across all PA tracts:", dollar(overall_median_income), "\n")

```


---

#### Step 3: Define Vulnerable Populations 

Identify census tracts with vulnerable populations based on TWO criteria:
1. Low median household income (choose an appropriate threshold)
2. Significant elderly population (choose an appropriate threshold)

**Your Task:**
```{r}
#### Step 3: Define Vulnerable Populations ----

income_threshold <- 50000
elderly_threshold <- 0.15

census_tracts_demo <- census_tracts_demo %>%
  mutate(
    pct_elderly =  pop_65_over/ total_pop
  )

vulnerable_tracts <- census_tracts_demo %>%
  filter(
    median_income < income_threshold &
    pct_elderly > elderly_threshold
  )

num_vulnerable <- nrow(vulnerable_tracts)
total_tracts <- nrow(census_tracts_demo)
pct_vulnerable <- (num_vulnerable / total_tracts) * 100

cat("Income threshold:", income_threshold, "\n")
cat("Elderly population threshold:", elderly_threshold * 100, "%\n")
cat("Number of vulnerable tracts:", num_vulnerable, "\n")
cat("Percentage of PA tracts considered vulnerable:", round(pct_vulnerable, 2), "%\n")

```

---

#### Step 4: Calculate Distance to Hospitals 

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**
```{r}
library(units)
# Transform to appropriate projected CRS
projected_crs <- 26918

vulnerable_tracts_proj <- st_transform(vulnerable_tracts, crs = projected_crs)
hospitals_proj <- st_transform(hospitals, crs = projected_crs)

tract_centroids <- st_centroid(vulnerable_tracts_proj)

dist_matrix <- st_distance(tract_centroids, hospitals_proj)
nearest_hospital_dist_m <- apply(dist_matrix, 1, min)

nearest_hospital_dist_miles <- as.numeric(nearest_hospital_dist_m) / 1609.34

vulnerable_tracts_proj <- vulnerable_tracts_proj %>%
  mutate(distance_to_hospital_miles = nearest_hospital_dist_miles)

avg_distance <- mean(vulnerable_tracts_proj$distance_to_hospital_miles, na.rm = TRUE)
max_distance <- max(vulnerable_tracts_proj$distance_to_hospital_miles, na.rm = TRUE)
num_over_15 <- sum(vulnerable_tracts_proj$distance_to_hospital_miles > 15, na.rm = TRUE)

cat("Average distance to nearest hospital (miles):", round(avg_distance, 2), "\n")
cat("Maximum distance to nearest hospital (miles):", round(max_distance, 2), "\n")
cat("Number of vulnerable tracts >15 miles from hospital:", num_over_15, "\n")

```

**Requirements:**
- Use an appropriate projected coordinate system for Pennsylvania
- Calculate distances in miles2
- Explain why you chose your projection

---

#### Step 5: Identify Underserved Areas 

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**
```{r}
# Create underserved variable

library(dplyr)

vulnerable_tracts_proj <- vulnerable_tracts_proj %>%
  mutate(
    underserved = distance_to_hospital_miles > 15
  )

num_underserved <- sum(vulnerable_tracts_proj$underserved, na.rm = TRUE)
total_vulnerable <- nrow(vulnerable_tracts_proj)
pct_underserved <- (num_underserved / total_vulnerable) * 100

cat("Number of underserved vulnerable tracts:", num_underserved, "\n")
cat("Percentage of vulnerable tracts that are underserved:", round(pct_underserved, 2), "%\n")

```

---

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**
```{r}
# Spatial join tracts to counties
#### Step 6: Aggregate to County Level ----

vulnerable_tracts_proj <- st_transform(vulnerable_tracts_proj, st_crs(pa_counties))

tracts_with_county <- st_join(vulnerable_tracts_proj, pa_counties, join = st_intersects, left = FALSE)

names(tracts_with_county)

county_summary <- tracts_with_county %>%
  st_drop_geometry() %>%
  group_by(NAMELSADCO) %>%
  summarize(
    num_vulnerable_tracts = n(),
    num_underserved_tracts = sum(underserved, na.rm = TRUE),
    pct_underserved = (num_underserved_tracts / num_vulnerable_tracts) * 100,
    avg_distance_miles = mean(distance_to_hospital_miles, na.rm = TRUE),
    total_vulnerable_pop = sum(total_pop, na.rm = TRUE)
  ) %>%
  arrange(desc(pct_underserved))

head(county_summary)

# Aggregate statistics by county

```

**Required county-level statistics:**
- Number of vulnerable tracts
- Number of underserved tracts  
- Percentage of vulnerable tracts that are underserved
- Average distance to nearest hospital for vulnerable tracts
- Total vulnerable population


**Questions to answer:**
- Which 5 counties have the highest percentage of underserved vulnerable tracts?
- Which counties have the most vulnerable people living far from hospitals?
- Are there any patterns in where underserved counties are located?
```{r}
top5_underserved <- county_summary %>%
  arrange(desc(pct_underserved)) %>%
  slice(1:5)
print(top5_underserved)

```
```{r}
top_vuln_far <- county_summary %>%
  arrange(desc(total_vulnerable_pop)) %>%
  slice(1:5)
print(top_vuln_far)

```
Most underserved counties tend to be located in rural northern and western Pennsylvania, where hospital density is lower and population centers are more dispersed. Urban counties (like Philadelphia or Allegheny) show fewer underserved tracts due to higher hospital accessibility.

---

#### Step 7: Create Summary Table 

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**
```{r}
# Create and format priority counties table
#### Step 7: Create Summary Table ----

library(knitr)
library(scales)

priority_metric <- "pct_underserved"

priority_counties <- county_summary %>%
  arrange(desc(!!sym(priority_metric))) %>%
  slice(1:10) %>%
  mutate(
    num_vulnerable_tracts = comma(num_vulnerable_tracts),
    num_underserved_tracts = comma(num_underserved_tracts),
    pct_underserved = percent(pct_underserved / 100, accuracy = 0.1),
    avg_distance_miles = round(avg_distance_miles, 1),
    total_vulnerable_pop = comma(total_vulnerable_pop)
  ) %>%
  rename(
    `County` = NAMELSADCO,
    `Vulnerable Tracts` = num_vulnerable_tracts,
    `Underserved Tracts` = num_underserved_tracts,
    `% Underserved` = pct_underserved,
    `Avg Distance (mi)` = avg_distance_miles,
    `Total Vulnerable Population` = total_vulnerable_pop
  )

kable(
  priority_counties,
  caption = "Table 1. Top 10 Pennsylvania Counties Prioritized for Healthcare Investment",
  align = "lccccc",
  format = "markdown"
)

```

**Requirements:**
- Use `knitr::kable()` or similar for formatting
- Include descriptive column names
- Format numbers appropriately (commas for population, percentages, etc.)
- Add an informative caption
- Sort by priority (you decide the metric)

---

## Part 2: Comprehensive Visualization 

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth 

Create a choropleth map showing healthcare access challenges at the county level.

```{r}
library(stringr)

county_summary <- county_summary %>%
  mutate(NAMELSADCO = str_replace(NAMELSADCO, " County$", ""))

```

```{r}
View(pa_counties)
View(county_summary)
```


**Your Task:**
```{r}
# Create county-level access map
#### Part 2: County-Level Choropleth ----

library(ggplot2)
library(sf)
library(scales)

pa_counties_proj <- st_transform(pa_counties, 26918)
hospitals_proj <- st_transform(hospitals, 26918)


county_summary_map <- pa_counties_proj %>%
  mutate(COUNTY_NAM = tolower(COUNTY_NAM)) %>%
  left_join(
    county_summary %>%
      mutate(NAMELSADCO = tolower(NAMELSADCO)),
    by = c("COUNTY_NAM" = "NAMELSADCO")
  )


ggplot() +
  geom_sf(
    data = county_summary_map,
    aes(fill = pct_underserved),
    color = "white",
    size = 0.2
  ) +

  scale_fill_gradient(
    low = "#DEEBF7",
    high = "#2171B5",
    name = "% Underserved Tracts",
    labels = function(x) paste0(round(x, 0), "%"),
    na.value = "grey90"
  ) +

  labs(
    title = "Healthcare Access Challenges Across Pennsylvania Counties",
    subtitle = "Percentage of vulnerable tracts that are more than 15 miles from the nearest hospital",
    caption = "Data sources: ACS 2022 (via tidycensus), Pennsylvania hospital data (Week 4 lecture)"
  ) +

  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 0.95, color = "gray40"),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9)
  )



```

**Requirements:**
- Fill counties by percentage of vulnerable tracts that are underserved
- Include hospital locations as points
- Use an appropriate color scheme
- Include clear title, subtitle, and caption
- Use `theme_void()` or similar clean theme
- Add a legend with formatted labels

---

### Map 2: Detailed Vulnerability Map 

Create a map highlighting underserved vulnerable tracts.

**Your Task:**
```{r}
# Create detailed tract-level map
#### Part 2: Map 2 - Detailed Vulnerability Map ----

pa_counties_proj <- st_transform(pa_counties, 26918)
hospitals_proj <- st_transform(hospitals, 26918)
vulnerable_tracts_proj <- st_transform(vulnerable_tracts_proj, 26918)

tract_map_data <- vulnerable_tracts_proj %>%
  mutate(
    tract_status = case_when(
      underserved ~ "Underserved Vulnerable Tract",
      TRUE ~ "Other Vulnerable Tract"
    )
  )

ggplot() +
  geom_sf(data = pa_counties_proj, fill = "gray95", color = "white", size = 0.3) +
  
  geom_sf(
    data = tract_map_data,
    aes(fill = tract_status),
    color = NA,
    alpha = 0.9
  ) +
  
  geom_sf(
    data = hospitals_proj,
    color = "red",
    size = 1.2,
    alpha = 0.8
  ) +
  
  scale_fill_manual(
    values = c(
      "Underserved Vulnerable Tract" = "#d73027",
      "Other Vulnerable Tract" = "#fee090"
    ),
    name = "Tract Type"
  ) +
  
  labs(
    title = "Underserved Vulnerable Tracts in Pennsylvania",
    subtitle = "Vulnerable tracts (low income & elderly) more than 15 miles from the nearest hospital are shown in red",
    caption = "Data: ACS 2022 (via tidycensus) and PA hospital data (Week 4 lecture)"
  ) +
  
  theme_void(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray20"),
    plot.caption = element_text(size = 9, color = "gray40", hjust = 0.95),
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9)
  )

```

**Requirements:**
- Show underserved vulnerable tracts in a contrasting color
- Include county boundaries for context
- Show hospital locations
- Use appropriate visual hierarchy (what should stand out?)
- Include informative title and subtitle

---

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**
```{r}
# Create distribution visualization
#### Chart: Distribution Analysis ----

library(ggplot2)
library(scales)

ggplot(vulnerable_tracts_proj, aes(x = distance_to_hospital_miles)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "#74c476",
    color = "white",
    alpha = 0.8
  ) +
  geom_density(color = "#006d2c", linewidth = 1) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", linewidth = 1) +
  labs(
    title = "Distribution of Distances to Nearest Hospital for Vulnerable Tracts",
    subtitle = "Dashed red line indicates the 15-mile underserved threshold",
    x = "Distance to Nearest Hospital (miles)",
    y = "Density",
    caption = "Data: ACS 2022 (via tidycensus) and PA hospital data (Week 4 lecture)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30"),
    plot.caption = element_text(size = 9, color = "gray40", hjust = 0.95),
    axis.title = element_text(face = "bold")
  )




```

**Suggested chart types:**
- Histogram or density plot of distances
- Box plot comparing distances across regions
- Bar chart of underserved tracts by county
- Scatter plot of distance vs. vulnerable population size

**Requirements:**
- Clear axes labels with units
- Appropriate title
- Professional formatting
- Brief interpretation (1-2 sentences as a caption or in text)

---

## Part 3: Bring Your Own Data Analysis 

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Challenge Options

Choose ONE of the following challenge exercises, or propose your own research question using OpenDataPhilly data (https://opendataphilly.org/datasets/). 

**Note these are just loose suggestions to spark ideas - follow or make your own as the data permits and as your ideas evolve. This analysis should include bringing in your own dataset, ensuring the projection/CRS of your layers align and are appropriate for the analysis (not lat/long or geodetic coordinate systems). The analysis portion should include some combination of spatial and attribute operations to answer a relatively straightforward question**

---

#### Education & Youth Services

**Option A: Educational Desert Analysis**
- **Data:** Schools, Libraries, Recreation Centers, Census tracts (child population)
- **Question:** "Which neighborhoods lack adequate educational infrastructure for children?"
- **Operations:** Buffer schools/libraries (0.5 mile walking distance), identify coverage gaps, overlay with child population density
- **Policy relevance:** School district planning, library placement, after-school program siting

**Option B: School Safety Zones**
- **Data:** Schools, Crime Incidents, Bike Network
- **Question:** "Are school zones safe for walking/biking, or are they crime hotspots?"
- **Operations:** Buffer schools (1000ft safety zone), spatial join with crime incidents, assess bike infrastructure coverage
- **Policy relevance:** Safe Routes to School programs, crossing guard placement

---

#### Environmental Justice

**Option C: Green Space Equity** 
- **Data:** Parks, Street Trees, Census tracts (race/income demographics)
- **Question:** "Do low-income and minority neighborhoods have equitable access to green space?"
- **Operations:** Buffer parks (10-minute walk = 0.5 mile), calculate tree canopy or park acreage per capita, compare by demographics
- **Policy relevance:** Climate resilience, environmental justice, urban forestry investment
---

#### Public Safety & Justice

**Option D: Crime & Community Resources**
- **Data:** Crime Incidents, Recreation Centers, Libraries, Street Lights
- **Question:** "Are high-crime areas underserved by community resources?"
- **Operations:** Aggregate crime counts to census tracts or neighborhoods, count community resources per area, spatial correlation analysis
- **Policy relevance:** Community investment, violence prevention strategies
---

#### Infrastructure & Services

**Option E: Polling Place Accessibility**
- **Data:** Polling Places, SEPTA stops, Census tracts (elderly population, disability rates)
- **Question:** "Are polling places accessible for elderly and disabled voters?"
- **Operations:** Buffer polling places and transit stops, identify vulnerable populations, find areas lacking access
- **Policy relevance:** Voting rights, election infrastructure, ADA compliance

---

#### Health & Wellness

**Option F: Recreation & Population Health**
- **Data:** Recreation Centers, Playgrounds, Parks, Census tracts (demographics)
- **Question:** "Is lack of recreation access associated with vulnerable populations?"
- **Operations:** Calculate recreation facilities per capita by neighborhood, buffer facilities for walking access, overlay with demographic indicators
- **Policy relevance:** Public health investment, recreation programming, obesity prevention

---

#### Emergency Services

**Option G: EMS Response Coverage**
- **Data:** Fire Stations, EMS stations, Population density, High-rise buildings
- **Question:** "Are population-dense areas adequately covered by emergency services?"
- **Operations:** Create service area buffers (5-minute drive = ~2 miles), assess population coverage, identify gaps in high-density areas
- **Policy relevance:** Emergency preparedness, station siting decisions

---

#### Arts & Culture

**Option H: Cultural Asset Distribution**
- **Data:** Public Art, Museums, Historic sites/markers, Neighborhoods
- **Question:** "Do all neighborhoods have equitable access to cultural amenities?"
- **Operations:** Count cultural assets per neighborhood, normalize by population, compare distribution across demographic groups
- **Policy relevance:** Cultural equity, tourism, quality of life, neighborhood identity

---

### Data Sources

**OpenDataPhilly:** https://opendataphilly.org/datasets/
- Most datasets available as GeoJSON, Shapefile, or CSV with coordinates
- Always check the Metadata for a data dictionary of the fields.

**Additional Sources:**
- **Pennsylvania Open Data:** https://data.pa.gov/
- **Census Bureau (via tidycensus):** Demographics, economic indicators, commute patterns
- **TIGER/Line (via tigris):** Geographic boundaries

### Recommended Starting Points

**If you're feeling confident:** Choose an advanced challenge with multiple data layers. 
**If you are a beginner, choose something more manageable that helps you understand the basics**


**If you have a different idea:** Propose your own question! Just make sure:
- You can access the spatial data
- You can perform at least 2 spatial operations

### Your Analysis

**Your Task:**

1. **Find and load additional data**
   - Document your data source
   - Check and standardize the CRS
   - Provide basic summary statistics

```{r}
# Load your additional dataset
base_path = "D:\\Desktop\\musa\\5080_policy\\portfolio-setup-xqdqc-hub\\assignments\\assignment_2\\data"

schools <- st_read(file.path(base_path, "Schools\\Schools.shp"))
crime <- st_read(file.path(base_path,"incidents_part1_part2\\incidents_part1_part2.shp"))
bike_net <- st_read(file.path(base_path,"Bike_Network\\Bike_Network.shp"))
```
```{r}
cat("Schools CRS:\n"); print(st_crs(schools))
cat("\nCrime Incidents CRS:\n"); print(st_crs(crime))
cat("\nBike Network CRS:\n"); print(st_crs(bike_net))
```
```{r}
crime <- st_transform(crime, 3857)
```

```{r}
cat("\nSchools:\n")
cat("Number of schools:", nrow(schools), "\n")
print(st_geometry_type(schools) %>% table())
print(summary(st_bbox(schools)))

cat("\nCrime Incidents:\n")
cat("Number of records:", nrow(crime), "\n")
print(st_geometry_type(crime) %>% table())
print(summary(st_bbox(crime)))

cat("\nBike Network:\n")
cat("Number of segments:", nrow(bike_net), "\n")
print(st_geometry_type(bike_net) %>% table())
print(summary(st_bbox(bike_net)))
```

**Questions to answer:**
- What dataset did you choose and why?
I have chosen School dataset because I'm interested in young children issue.

- What is the data source and date?
Crime Incidents is in 2024, and I also have school dataset and bike network dataset.

- How many features does it contain?
Number of schools is 495, number of crime records is 160388, number of bike network segments is 5225.

- What CRS is it in? Did you need to transform it?
The CRS of schools and bike network data is EPSG 3857, the CRS of crime data is EPSG 4326. So I have got crime data transformed.

---

2. **Pose a research question**

Write a clear research statement that your analysis will answer.

This analysis investigates whether school zones in Philadelphia are safe environments, by examining the number of nearby crime incidents.

---

3. **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):**
- Buffers
- Spatial joins
- Spatial filtering with predicates
- Distance calculations
- Intersections or unions
- Point-in-polygon aggregation

**Your Task:**
```{r}
# Your spatial analysis
school_buffers <- st_buffer(schools, dist = 1000)

crime_join <- st_join(crime, school_buffers, join = st_within, left = FALSE)
```

```{r}
View(crime_join)
```


```{r}
school_crime_count <- crime_join %>%
  st_drop_geometry() %>%
  group_by(school_nam) %>%
  summarize(crime_count = n())

schools_with_crime <- schools %>%
  left_join(school_crime_count, by = "school_nam") %>%
  mutate(crime_count = replace_na(crime_count, 0))
```

```{r}
head(schools_with_crime)
```

```{r}
bike_in_buffers <- st_intersection(bike_net, school_buffers)

bike_coverage <- bike_in_buffers %>%
  st_drop_geometry() %>%
  group_by(school_nam) %>%
  summarize(bike_miles = sum(st_length(bike_in_buffers) / 5280))

schools_with_bike <- schools_with_crime %>%
  left_join(bike_coverage, by = c("school_nam")) %>%
  mutate(bike_miles = replace_na(bike_miles))

head(schools_with_bike)
```


```{r}
summary_tbl <- schools_with_bike %>%
  st_drop_geometry() %>%
  summarize(
    total_schools = n(),
    avg_crime = mean(crime_count, na.rm = TRUE),
    avg_bike = mean(bike_miles, na.rm = TRUE)
  )

print(summary_tbl)
```



```{r}
ggplot(data = schools_with_crime) +
  geom_sf(aes(color = crime_count), size = 2) +
  scale_color_viridis_c(
    option = "plasma",
    trans = "sqrt",
    name = "Crime Count"
  ) +
  labs(
    title = "Crime Incidents Within 1000ft of Philadelphia Schools",
    subtitle = "Each point represents a school, colored by number of nearby crime incidents",
    caption = "Data: OpenDataPhilly | Analysis: MUSA 5080 Assignment 3"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "gray30"),
    legend.position = "right"
  )

```


```{r}
ggplot(schools_with_crime, aes(x = crime_count)) +
  geom_histogram(
    bins = 30,
    fill = "pink",
    color = "white",
    alpha = 0.8
  ) +
  labs(
    title = "Distribution of Crime Incidents Within 1000ft of Schools",
    subtitle = "Each bar represents the number of schools by nearby crime count",
    x = "Number of Crime Incidents (within 1000ft)",
    y = "Number of Schools",
    caption = "Data: OpenDataPhilly | Analysis: MUSA 5080 Assignment 3"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "gray30"),
    plot.caption = element_text(size = 8, color = "gray50", hjust = 0.95)
  )

```


**Analysis requirements:**
- Clear code comments explaining each step
- Appropriate CRS transformations
- Summary statistics or counts
- At least one map showing your findings
- Brief interpretation of results (3-5 sentences)

**Your interpretation:**
The majority of Philadelphia schools experience a moderate number of nearby crime incidents, with most falling below 2,000 cases within a 1,000-ft radius. However, a few schools stand out as clear hotspots, with significantly higher counts exceeding 6,000 incidents, concentrated primarily in central and southern parts of the city. The histogram confirms this right-skewed distribution, where a small proportion of schools account for a disproportionately large share of nearby crimes. These findings suggest that while most school zones remain relatively safe, targeted safety interventions and “Safe Routes to School” programs should prioritize the high-crime areas identified in the analysis.


## Finally - A few comments about your incorporation of feedback!
Take a few moments to clean up your markdown document and then write a line or two or three about how you may have incorporated feedback that you recieved after your first assignment. 

---

## Submission Requirements

**What to submit:**

1. **Rendered HTML document posted to your course portfolio** with all code, outputs, maps, and text
   - Use `embed-resources: true` in YAML so it's a single file
   - All code should run without errors
   - All maps and charts should display correctly

**File naming:** `LastName_FirstName_Assignment2.html` and `LastName_FirstName_Assignment2.qmd`

---

